---
title: "Sampling design"
author: "Bharti Dharapuram"
date: "2023-09-28"
output:
  html_document: default
  pdf_document: default
---

# Objectives

In this tutorial we will use a polygon of Silent Valley National Park to:

1. draw nested sampling grids and choose random sampling locations
2. choose random sampling locations stratified by environmental raster

We will primarily be working with the packages `sf` and `terra` to handle spatial data, use the package `dplyr` for data manipulation and `ggplot2` for visualization with a little help from `tidyterra` and `ggpubr`.

# Setup
```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

# Call the required R libraries

```{r}
library(sf)
library(terra)
library(dplyr)
library(ggplot2)
library(tidyterra)
library(ggpubr)
```

---

# 1. Input spatial data

## 1.1. Protected areas of Kerala and subset data

We will load the protected area list of Kerala and keep the polygon corresponding to Silent Valley National Park.

```{r}
# load shape files of protected areas of Kerala from disk
pa<-st_read('data/kerala_pa_6Jun22/kerala_pa_6Jun22.shp')
head(pa) #look at the organization of the data in this polygon file
pa$pa #look at the list of names of protected areas, so that we can pull out the data we need
```

We want to keep the polygon corresponding to Silent Valley National Park from the list. Filtering the data based on the protected area name, and plotting to see the results.

```{r}
sv<-pa %>%
  filter(pa=="Silent Valley National Park")
head(sv)
plot(st_geometry(sv))
```

## 1.2. Environmental variables from WorldClim - precipitation seasonality and elevation
```{r}
prec_seas<-terra::rast("data/wc2.1_30s_bio_15.tif")
elev<-terra::rast("data/wc2.1_30s_elev.tif")
```

Crop the above rasters to the extent of Silent Valley National Park.

```{r}
#sv_ext<-st_bbox(sv) #obtaining the extent of the Silent Valley National Park polygon
elev<-terra::crop(elev, sv, mask=T) #cropping the elevation raster to the Silent Valley extent
prec_seas<-terra::crop(prec_seas, sv, mask=T) #cropping the precipitation seasonality raster to the Silent Valley extent
```

Plot the the polygon over the cropped rasters.

```{r}
elev.fill<-scale_fill_gradientn(name="Elevation (m)", colours=terrain.colors(10), na.value="transparent") #choose the fill palette for elevation data
prec_seas.fill<-scale_fill_gradient(name="Precipitation CV", low="#f7fbff", high="#08519c", na.value="transparent") #choose the fill palette for precipitation seasonality data
my.xscale<-scale_x_continuous(breaks=seq(76.38, 76.48, 0.05)) #define longitudinal bins
my.yscale<-scale_y_continuous(breaks=seq(11, 11.25, 0.05)) #define latitudinal bins

plt_elev<-ggplot() +
  geom_spatraster(data=elev) + #plot raster data
  geom_sf(data=sv, fill=NA, col="black", linewidth=0.4) + #plot the polygon corresponding to Silent Valley National Park
  elev.fill +
  my.xscale +
  my.yscale +
  theme(legend.position="bottom")

plt_prec_seas<-ggplot() +
  geom_spatraster(data=prec_seas) + #plot raster data
  geom_sf(data=sv, fill=NA, col="black", linewidth=0.4) + #plot the polygon corresponding to Silent Valley National Park
  prec_seas.fill +
  my.xscale +
  my.yscale +
  theme(legend.position="bottom")

ggarrange(plt_elev, plt_prec_seas) #arranging multiple ggplots on the same page
```

# 2. Overlay sampling grid over PA polygon and draw random sampling locations

Lets create 2.5 x 2.5 km grids over Silent Valley National Park. For this, we would first need to project the polygon vector so that the unit of measurement is in distance units.

```{r}
sv_proj<-st_transform(sv, crs="EPSG:32643") #to reproject the polygon, we use the projection EPSG: 32643 corresponding to UTM zone 43 in the northern hemisphere

sv_gridl<-st_make_grid(sv_proj, cellsize=c(2500, 2500)) %>%
  st_as_sf() #create a grid using the projected polygon as input and specifying the cell size (in meters)

sv_grids<-st_make_grid(sv_proj, cellsize=c(500, 500)) %>%
  st_as_sf() #create a nested grid that is 500 x 500 m in size 

#make a simple plot to view the sampling grid
plot(st_geometry(sv_proj), col="blue")
plot(sv_grids, add=TRUE)
plot(sv_gridl, border="red", add=TRUE)

#retaining those grids that intersect with the polygon area and converting it into an sf object
sv_gridl<-sv_gridl %>%
  st_filter(sv_proj, .predicates=st_intersects)

sv_grids<-sv_grids %>%
  st_filter(sv_proj, .predicates=st_intersects)

st_coordinates(sv_gridl)

#plotting the above objects
plot(st_geometry(sv_proj), col="blue")
plot(sv_grids, add=TRUE)
plot(sv_gridl, border="red", add=TRUE)

#sample random polygons from the grid
set.seed(3432)
l<-sample(c(1:nrow(sv_gridl)), 6) #sample random row numbers
sv_gridl_sample<-sv_gridl[l,] #subset random rows from the sf dataframe
sv_gridl_sample$large.id<-c(seq(1:nrow(sv_gridl_sample))) #create a column of polygon ids

#plotting the selected polygons
plot(st_geometry(sv_proj), col="blue")
plot(sv_gridl_sample, col="grey", add=TRUE)
plot(sv_grids, add=TRUE)
plot(sv_gridl, border="red", add=TRUE)

#shrink the larger grid cells by a little
sv_gridl_sample2<-st_buffer(sv_gridl_sample, -100) #if we do not do this, the adjacent smaller grid cells adjacent to the large grid cell would be considered to fall within the large grid cell
sv_grids_sample<-sv_grids %>% 
  st_filter(sv_gridl_sample2, .predicates=st_intersects) #filter the small grid cells that fall within each of the randomly sampled larger grid cells
sv_grids_sample$small.id<-seq(1:nrow(sv_grids_sample)) #create a column of polygon ids

#join information of sampled large grids with the small grids
sv_gridsl<-st_join(sv_grids_sample, sv_gridl_sample) #add information of the large grid cell ids to the small grid cells

#sample five random grids within larger grids
sv_gridsl_sample<-sv_gridsl %>%
  group_by(large.id) %>%
  slice_sample(n=5) #within a group sample random rows

#obtain centroids of the selected small grid cells
sv_sample_pts<-st_as_sf(st_centroid(st_geometry(sv_gridsl_sample)))

#plot the randomly sampled plots
plot(st_geometry(sv_proj), col="blue")
plot(sv_gridl_sample, border="red", col="grey", add=TRUE)
plot(st_geometry(sv_gridsl_sample), add=TRUE)
plot(sv_sample_pts, pch=19, cex=0.75, add=TRUE)
```

Write the centroids of the chosen small polygons to disk.

```{r}
#create a directory where the output needs to be stored
dir.create("results")
st_write(sv_sample_pts, "results/sampling_locations.shp")
```

# 3. Select sampling locations based on stratified sampling over raster values

## 3.1 Look at the distribution of elevation data

Divide the raster values into quartiles and define strata based on it. Create a new raster where the group numbers are assigned as values and choose random samples based from each of the elevation strata

```{r}
elev_qtile<-values(elev) %>% #pull out the values of the raster
  as.data.frame() %>% #convert into a dataframe so that new columns can be added
  mutate(cell=seq(1:nrow(values(elev)))) %>% #add a new column of cell numbers corresponding to these values
  mutate(quartile=ntile(wc2.1_30s_elev, 4)) #create quartile

elev2<-elev #make a copy of the elevation raster so that values can be reassigned
elev2[elev_qtile$cell]<-elev_qtile$quartile #based on the cell index replace elevation values with the quartile classes

#quick plot to look at the reassignment of values
par(mfrow=c(1,2))
plot(elev)
plot(sv, col=NA, add=TRUE)
plot(elev2)
plot(sv, col=NA, add=TRUE)
```

Plot the potential sampling locations drawn from the different sampling techniques.

```{r}
set.seed(2341)

#sample random points from each of the quartile groups
sampl_str<-spatSample(elev2, size=4, method="stratified", as.points=TRUE, na.rm=TRUE)

#create a random selection of points for comparison
sampl_rnd<-spatSample(elev, size=16, as.points=TRUE, na.rm=TRUE)

#plot the each of these sampled locations (random and) over the raster
par(mfrow=c(2,2))
plot(elev, main="Random sampling: Elevation")
plot(sv, col=NA, add=TRUE)
points(sampl_rnd)

plot(elev, main="Stratified sampling: Elevation")
plot(sv, col=NA, add=TRUE)
points(sampl_str)

plot(elev2, main="Random sampling: Quantile classes")
plot(sv, col=NA, add=TRUE)
points(sampl_rnd)

plot(elev2, main="Stratified sampling: Quantile classes")
plot(sv, col=NA, add=TRUE)
points(sampl_str)
```

---